Interactive Graphics for Visually Diagnosing Forest Classifiers in R

### 2022-10-18 COST_submission_18_Oct22/PPforestVizupdate.Rnw
update repo

\institute{Natalia da Silva \at
              Instituto de Estadística (IESTA), Universidad de la República\\
              \email{natalia.dasilva@fcea.edu.uy}
           \and
          Dianne Cook \at 
          Department of Econometrics and Business Statistics, Monash University\\
          \email{dicook@monash.edu}\and
          Eun-Kyung Lee \at
          Department of Statistics, Ewha Womans University\\
          \email{lee.eunk@gmail.com}
}

>>>>

\institute{Corresponding author: Natalia da Silva \at
              Instituto de Estadística (IESTA), Universidad de la República, Montevideo, Uruguay.\\
              \email{natalia.dasilva@fcea.edu.uy}\\
              ORCID:0000-0002-6031-7451\\
           \and
          Dianne Cook \at 
          Department of Econometrics and Business Statistics, Monash University, Melbourne, Australia.\\
          \email{dicook@monash.edu}\\
          ORCID:0000-0002-3813-7155\\
          \and
          Eun-Kyung Lee \at
          Department of Statistics, Ewha Womans University, Seoul, South Korea.\\
          \email{lee.eunk@gmail.com}
}

### 2022-10-18 COST_submission_18_Oct22/PPforestVizupdate.tex
update repo

\begin{figure}[!t]
\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
\includegraphics[width=\maxwidth]{boundss-1} 
\end{knitrout}
 \vspace*{-0.3cm}
 \caption{Decision boundaries for PPtree algorithm on 2D simulated data. The partitions generated by PPtree algorithm are oblique to the axis, incorporating the association between the two variables.\label{bounds}}
\end{figure}
Two data set are used in this paper, crab~\citep{CM74} and fishcatch data~\citep{fishcatch}. To illustrate the mapping of diagnostics to visualization crab data are used because of its simplicity. The second data set is introduced to illustrate the interactive web app which combines the visualizations introduced in this paper and present a more complex problem to explore within the application.

@@

\includegraphics[width=\maxwidth]{trcomb-2} 
\end{knitrout}
\caption{Visualizing the PPtree model of the crab data. The tree has three nodes (top). The density plots show the data projections at each node, colored by group (middle). The dashed vertical red line indicates the split value of each node. Node 1 separates the sexes. Node 2 separate males from species and node 3 separate females from species. Mosaic plots of the confusion table for each split (bottom). Node 1 shows the clear split of sexes, with a small number of misclassifications. Node 2 where BM are separated from OM with small number of misclassifications. Node 3 where BF are separated from OF, almost perfect classification.\label{pptreeviz1}}
\end{figure}

@@

Figure \ref{impotree} shows the absolute projection coefficient of the top three nodes for all the trees in a forest model. This information is displayed by a side-by-side jittered dot plot. The red dots correspond to the absolute coefficient values for the tree model of Figure \ref{pptreeviz1}. The forest was built using random samples of two variables for each node, hence there are two coefficients for each node. At node 1 uses CW and RW with similar contribution. The scatterplot at right shows these two variables and the resulting boundary between groups that this would produce. Node 2 uses BD and CL where BD contributes the most to the separation. The plot in the right shows the boundary that is induced. Node 3 uses CW and FL with an even contribution by the two variables.
For each tree in the forest, different decision rules are defined; the resulting boundaries on the previous plots are based on Rule 1 $= \frac{m_1}{2} + \frac{m_2}{2} $, where $m_1$ and $m_2$ are the mean of the left and right groups at each node.

@@

\begin{figure}[hbpt]
\centering
\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
\includegraphics[width=\maxwidth,trim=0cm 2cm 0cm 2cm, clip]{ternary-1} 
\end{knitrout}
\caption{Generalized ternary plot ((G-1) dimensional simplex, in this case a tetrahedron) representation of the vote matrix for four classes. The tetrahedron is shown pairwise. Each point corresponds to one observation and color is the true class. This is close but not a perfect classification since the colors are concentrated in the corners and there are some mixed colors in each corner.}
\label{tetra}
\end{figure}

@@

Figure \ref{tab1} shows the arrangement of plots in this first tab (the case level tab).

This selection of plots enables aspects of the model, relating to performance for individual cases, to be examined in the data space. The data plot is an essential element following the {\em model-in-the-data-space} philosophy of \citet{wickham2015visualizing}. The choice was made to use a parallel coordinate plot because it provides a space-efficient display of the data. Alternatives include the tour, a dynamic plot, or a scatterplot matrix. Theoretically, either of these could be substituted or added. 
Parallel coordinate plots are useful to visualize multivariate numerical data. As is shown in the first plot in Figure \ref{tab1}, in a PCP each variable is placed equidistant and perpendicular to the x-axis, on the y-axis the values for each variable are standardized to easily compare them. Each observation is represented as a line connected across  all the variables. In a PCP many variables can be compared together and seeing the relationships between them, identify and characterize groups of observations. The features can be ordered so that there are not too many lines intersecting resulting in an unreadable chart \citep{inselberg2009parallel}. 
Two alternatives can be selected in shiny to draw the parallel coordinate plot: parallel or enhanced. Parallel draws the classic PCP and enhanced draws a modified version where variables are repeated \citep{hurley2011eulerian}. Because reading a PCP is really only possible for neighboring variables, the variables are repeated so that all variable pairs are shown next to each other once.

\begin{figure}[hbpt]
\centering
\includegraphics[width=1\linewidth]{fish11.png}
\caption{Entry page for the web app, focusing on examining cases, in terms of similarity and uncertainty. The top plot shows the data, the remaining plots show similarity of cases and uncertainty in predictions. All of the plots are linked so that selecting elements of one will generate changes in the others. \label{tab1}}
\end{figure}

@@

\begin{figure}[hbpt]
\centering
\includegraphics[width=1\linewidth]{tab1com2.png}
\caption{Entry page for the web app, focusing on examining cases, for selected points.  \label{tab1comp}}
\end{figure}

@@

The selected tree in Figure \ref{tab2} has very small OOB error shown as red point in the boxplot. Two variables were use in each node partition, the first four nodes were selected and shown in the importance variable plot, density plot and mosaic plot. In the importance variable plot \verb#height# and \verb#lenght1# were used in the first node partition being \verb#height# the important variable to separate Bream and Parkki (Green and orange in density and mosaic plots) from the rest. In the second node \verb#length1# and \verb#length2# were both evenly important  to separate Bream from Parkki. A similar analysis can be done for other nodes. This tab allow us to understand individual trees in the forest and analyzed which variables are important for the node partition related with the OOB error involved in that specific tree.

\begin{figure}[hbpt]
\centering
\includegraphics[width=1\linewidth]{tab2comp.png}
\caption{The individual model tab in the web app. The OOB error is displayed in a boxplot  including jittered points for all trees in the forest. Variable importance is displayed as jittered dot plots for selected nodes of all trees. The boxplot is linked to a display of the PPTree, the variable importance plot, and a density plot of the data projections showing splits at each selected nodes and confusion tables as mosaic plots. Clicking a point in the OOB error boxplot triggers various updates: each of the importance values for the same tree are highlighted (red), the tree that this corresponds to is drawn, the error for the tree is shown on the boxplot (in red), and the density plot and the mosaic plot are also updated. Also specific nodes can be selected to update the plots.  \label{tab2}}
\end{figure}

@@

\begin{figure}[hbpt]
\centering
\includegraphics[width=1\linewidth]{fishT2.pdf}
\caption{Schematic diagram illustrating the interactivity in and between plots for model level exploration panel of the web app. The boxplot and the dotplot of variable importance has click selection, which invokes changes of the corresponding displays. Selecting a point in the boxplot update the node tree information (Shiny Gui) and update all the plots for the three first nodes, also nodes can be selected and update the corresponding plots. Finally selecting a point in the importance variable plot makes a change in the data, which propagates the importance values for other variables in this tree to be highlighted (red), draws the tree, highlights the error value of the tree, and shows the projections and confusion matrix for the three top nodes.}
\label{tab2diag}
\end{figure}
\newpage

@@

The  ROC is used to summarize the trade-off between sensitivity and specificity. The plot shows the sensitivity and specificity when a parameter of the classifier is varied \citep{trevor2011elements}. The specificity and sensitivity were computed with the \verb#pROC# package.
If more than two classes are available a multi-class ROC analysis is needed. Several solutions have been proposed for multi-class ROC. Some of the proposed reduced the multi-class problem to a set of binary problems. The approach used for a multi-class ROC analysis in this article is called one-against-all ~\citep{allwein2000reducing}.

\begin{figure}[hbpt]
\includegraphics[width=.8\linewidth]{fish31.png}
\caption{Performance comparison tab of the web app. ROC curves displaying sensitivity against specificity for the classes are shown, along with the OOB error by the number of trees used to build the forest, and overall variable importance. Displays are shown for the PPF and RF, for comparison purposes. Users can select a class to focus on, using the text entry box. \label{tab3}}
\end{figure}
\newpage

>>>>

\begin{figure}[!t]
\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
\includegraphics[width=\maxwidth]{boundss-1} 
\end{knitrout}
 \vspace*{-0.3cm}
 \caption{Decision boundaries for PPtree algorithm on 2D simulated data. The partitions generated by PPtree algorithm are oblique to the axis, incorporating the association between the two variables.\label{bounds}}
\end{figure}
Two data sets are used in this paper, crab~\citep{CM74} and fishcatch data~\citep{fishcatch}. To illustrate the mapping of diagnostics to visualization crab data are used because of its simplicity. The second data set is introduced to illustrate the interactive web app which combines the visualizations introduced in this paper and present a more complex problem to explore within the application.

@@

\includegraphics[width=\maxwidth]{trcomb-2} 
\end{knitrout}
\caption{Visualizing the PPtree model of the crab data. The tree has three nodes (top). The density plots show the data projections at each node, colored by group (middle). The dashed vertical red line indicates the split value of each node. Node 1 separates the sexes. Node 2 separates males from species and node 3 separates females from species. Mosaic plots of the confusion table for each split (bottom). Node 1 shows the clear split of sexes, with a small number of misclassifications. Node 2 where BM are separated from OM with a small number of misclassifications. Node 3 where BF are separated from OF, almost perfect classification.\label{pptreeviz1}}
\end{figure}

@@

Figure \ref{impotree} shows the absolute projection coefficient of the top three nodes for all the trees in a forest model. This information is displayed by a side-by-side jittered dot plot. The red dots correspond to the absolute coefficient values for the tree model of Figure \ref{pptreeviz1}. The forest was built using random samples of two variables for each node, hence there are two coefficients for each node. At node 1 uses CW and RW with similar contribution. The scatterplot at right shows these two variables and the resulting boundary between groups that this would produce. Node 2 uses BD and CL where BD contributes the most to the separation. The plot in the right shows the boundary that is induced. Node 3 uses CW and FL with an even contribution by the two variables. Alternative visualizations for the importance variable plot can be considered, for example violin plots or boxplots however, the side-by-side jittered dot plot was preferred since presents individual details on importance measure distribution.

For each tree in the forest, different decision rules are defined; the resulting boundaries on the previous plots are based on Rule 1 $= \frac{m_1}{2} + \frac{m_2}{2} $, where $m_1$ and $m_2$ are the mean of the left and right groups at each node.

@@

\begin{figure}[hbpt]
\centering
\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
\includegraphics[width=\maxwidth]{ternary-1} 
\end{knitrout}
\caption{Generalized ternary plot ((G-1) dimensional simplex, in this case a tetrahedron) representation of the vote matrix for four classes. The tetrahedron is shown pairwise. Each point corresponds to one observation and color is the true class. This is close but not a perfect classification since the colors are concentrated in the corners and there are some mixed colors in each corner.}
\label{tetra}
\end{figure}

@@

Figure \ref{tab1} shows the arrangement of plots in this first tab (the case level tab). This selection of plots enables aspects of the model, relating to performance for individual cases, to be examined in the data space. The data plot is an essential element following the {\em model-in-the-data-space} philosophy of \citet{wickham2015visualizing}. The choice was made to use a parallel coordinate plot because it provides a space-efficient display of the data. Alternatives include the tour, a dynamic plot, or a scatterplot matrix. Theoretically, either of these could be substituted or added.
Parallel coordinate plots are useful to visualize multivariate numerical data. As is shown in the first plot in Figure \ref{tab1}, in a PCP each variable is placed equidistant and perpendicular to the x-axis, on the y-axis the values for each variable are standardized to easily compare them. Each observation is represented as a line connected across  all the variables. In a PCP many variables can be compared together and seeing the relationships between them, identify and characterize groups of observations. The features can be ordered so that there are not too many lines intersecting resulting in an unreadable chart \citep{inselberg2009parallel}.
Two alternatives can be selected in shiny to draw the parallel coordinate plot: parallel or enhanced. Parallel draws the classic PCP and enhanced draws a modified version where variables are repeated \citep{hurley2011eulerian}. Because reading a PCP is really only possible for neighboring variables, the variables are repeated so that all variable pairs are shown next to each other once.

\begin{figure}[hbpt]
\centering
\includegraphics[width=1.05\linewidth]{fish11.png}
\caption{Entry page for the web app, focusing on examining cases, in terms of similarity and uncertainty. The top plot shows the data, the remaining plots show similarity of cases and uncertainty in predictions. All of the plots are linked so that selecting elements of one will generate changes in the others. \label{tab1}}
\end{figure}

@@

\begin{figure}[hbpt]
\centering
\includegraphics[width=1.1\linewidth]{tab1com2.png}
\caption{Entry page for the web app, focusing on examining cases, for selected points.  \label{tab1comp}}
\end{figure}

@@

The selected tree in Figure \ref{tab2} has a very small OOB error shown as a red point in the boxplot. Two variables were use in each node partition, the first four nodes were selected and shown in the importance variable plot, density plot and mosaic plot. In the importance variable plot \verb#height# and \verb#lenght1# were used in the first node partition being \verb#height# the important variable to separate Bream and Parkki (Green and orange in density and mosaic plots) from the rest. In the second node \verb#length1# and \verb#length2# were both evenly important  to separate Bream from Parkki. A similar analysis can be done for other nodes. This tab allows us to understand individual trees in the forest and analyze which variables are important for the node partition related to the OOB error involved in that specific tree.

\begin{figure}[hbpt]
\centering
\includegraphics[width=1.1\linewidth]{tab2comp.png}
\caption{The individual model tab in the web app. The OOB error is displayed in a boxplot  including jittered points for all trees in the forest. Variable importance is displayed as jittered dot plots for selected nodes of all trees. The boxplot is linked to a display of the PPTree, the variable importance plot, and a density plot of the data projections showing splits at each selected node and confusion tables as mosaic plots. Clicking a point in the OOB error boxplot triggers various updates: each of the importance values for the same tree are highlighted (red), the tree that this corresponds to is drawn, the error for the tree is shown on the boxplot (in red), and the density plot and the mosaic plot are also updated. Also specific nodes can be selected to update the plots.  \label{tab2}}
\end{figure}

@@

\begin{figure}[hbpt]
\centering
\includegraphics[width=1.05\linewidth]{fishT2.pdf}
\caption{Schematic diagram illustrating the interactivity in and between plots for model level exploration panel of the web app. The boxplot and the dotplot of variable importance has click selection, which invokes changes of the corresponding displays. Selecting a point in the boxplot update the node tree information (Shiny Gui) and update all the plots for the three first nodes, also nodes can be selected and update the corresponding plots. Finally selecting a point in the importance variable plot makes a change in the data, which propagates the importance values for other variables in this tree to be highlighted (red), draws the tree, highlights the error value of the tree, and shows the projections and confusion matrix for the three top nodes.}
\label{tab2diag}
\end{figure}
\newpage

@@

 The  ROC is used to summarize the trade-off between sensitivity and specificity. The plot shows the sensitivity and specificity when a parameter of the classifier is varied \citep{trevor2011elements}. The specificity and sensitivity were computed with the \verb#pROC# package.
 If more than two classes are available a multi-class ROC analysis is needed. Several solutions have been proposed for multi-class ROC. Some of the proposed reduced the multi-class problem to a set of binary problems. 
The approach used for a multi-class ROC analysis in this article is called one-against-all \citep{allwein2000reducing}.

\begin{figure}[hbpt]
\includegraphics[width=1\linewidth]{fish31.png}
\caption{Performance comparison tab of the web app. ROC curves displaying sensitivity against specificity for the classes are shown, along with the OOB error by the number of trees used to build the forest, and overall variable importance. Displays are shown for the PPF and RF, for comparison purposes. Users can select a class to focus on, using the text entry box. \label{tab3}}
\end{figure}
\newpage
